Follow-up Samples

### 2016-12-22 submission/manuscript_outline_20161024.Rmd
rewrite methods order include an analysis overview section

***Analysis Overview***
This

***Selection of Important OTUs***
In order to assess which variables were most central to all the models we counted the number of times a variable was present in the top 10% of mean decrease in accuracy (MDA) for each different 80/20 split model and then filtered this list to variables that were only present more than 50% of the time.  This final collated list of variables was what was considered the most important for the lesion prediction model.

A second model using the same types of conditions was used to identify OTUs that could predict initial and follow up samples.  These OTUs were considered the most important for predicting initial and follow up samples.  This new model important OTUs were then compared to the overall lesion model important OTUs and the resulting common OTUs to both models were selected to test for changes between initial and follow up using a paired wilcoxson rank sum test.

***Investigation of Important CRC Bacteria***
In order to investigate how specific bacteria that have been previously associated with CRC with selected OTUs that taxonomically classified to Fusobacterium Nucleatum, Parvimonas Micra, Peptostreptococcus Assacharolytica, and Porphyromonas Stomatis.  Specifically we wanted to observe if there were any differences based on whether the individual had an adenoma or carcinoma that may have been missed with our lesion versus normal model.

@@

***Reproducible methods.***
A detailed and reproducible description of how the data were processed and analyzed can be found at https://github.com/SchlossLab/Baxter_followUps_2016.

>>>>

***Selection of Important OTUs***
In order to assess which variables were most central to all the models we counted the number of times a variable was present in the top 10% of mean decrease in accuracy (MDA) for each different 80/20 split model and then filtered this list to variables that were only present more than 50% of the time.  This final collated list of variables was what was considered the most important for the lesion or initial follow up models.

@@

***Analysis Overview***
Differences in FIT between initial and follow ups for either adenoma or carcinoma were investigate.  Next, initial and follow up samples were analyzed for differences in alpha and beta diversity.  All OTUs used in the lesion model were also analyzed using a paired wilcoxson test.  The lesion model was then tested for accuracy in prediction and whether it reduced the positive probability of lesion after surgery.  The most important OTUs for this were used to build an updated model and this reduced feature model was assessed for it's similarity to the original model.  We then used the initial follow up model to assess whether this model could classify samples better then the lesion model.  The most important OTUs were then identified from this model and used to create a reduced feature initial follow up model.  This reduced feature model, as was done with the lesion model, was compared to the full model for loss of accuracy.  Finally, in order to investigate the relative abundance of specific bacteria, that have been previously associated with CRC, we selected OTUs that taxonomically classified to Fusobacterium Nucleatum, Parvimonas Micra, Peptostreptococcus Assacharolytica, and Porphyromonas Stomatis.  Specifically, we wanted to test if there were any differences based on whether the individual had an adenoma or carcinoma.

***Reproducible methods.***
A detailed and reproducible description of how the data were processed and analyzed can be found at https://github.com/SchlossLab/Sze_followUps_2017.

### 2016-12-22 submission/manuscript_outline_20161024.tex
rewrite methods order include an analysis overview section

\textbf{\emph{Study Design and Patient Sampling}} The sampling and
design of the study was similar to that reported in Baxter, et al
(\textbf{insert citation}). In brief, study exclusion involved those who
had already undergone surgery, radiation, or chemotherapy, had
colorectal cancer before a baseline stool sample could be obtained, had
IBD, a known hereditary non-polyposis colorectal cancer, or Familial
adenomatous polyposis. Samples used to build the model used for
prediction were collected either prior to a colonoscopy or between 1 - 2
weeks after. The bacterial microbiome has been shown to nomralize within
this time period (\textbf{insert citation}). Kept apart from this
training set were a total of 67 individuals that not only had a sample
as described previoulsy but also a follow up sample between 188 - 546
days after surgery and treatment had been completed. This study was
approved by the University of Michigan Institutional Review Board. All
study participants provided informed consent and the study itself
conformed to the guidelines set out by the Helsinki Declaration.

\textbf{\emph{Fecal Immunochemical Test and 16S rRNA Gene Sequencing}}
FIT was analyzed as previously published using both OC FIT-CHEK and
OC-Auto Micro 80 automated system (Polymedco Inc.) (\textbf{insert
citation}). 16S rRNA gene sequencing was completed as previously
described by Kozich, et al. (\textbf{insert citation}). In brief, DNA
extraction used the 96 well Soil DNA isolation kit (MO BIO Laboratories)
and an epMotion 5075 automated pipetting system (Eppendorf). The V4
variable region was amplified and the resulting product was split
between three sequencing runs with control, adenoma, and carcinoma
evenly represented on each run. Each group was randomly assigned to
avoid biases based on sample collection location.

@@

\textbf{\emph{Model Creation}} The Random Forest (\textbf{insert
citation}) algorithm was used to create the model used for prediction of
lesion (adenoma or carcinoma) for the 67 individuals with follow up
samples. The model included data on FIT, the bacterial microbiome, sex,
age, Body Mass Index (BMI), whether the indivdiual was caucasion or not,
history of cancer, and family history of cancer. Non-binary data was
checked for near zero variance and auto correlation. Data columns that
had near zero variance were removed. Columns that were correlated with
each other over a Spearman correlation coefficient of 0.75 had one of
the two columns removed. This pre-processing was performed with the R
package caret (v6.0.73). Optimiation of the mtry hyperparameter as well
as data on the best, middle, and worse performance of the model involved
taking the 490 samples and making 100 80/20 (train/test) splits in the
data where control and lesion were equally represented in the 80 and 20
split, respectively. This 80\% portion was then split again into an
80/20 split, and run through 20 repeated 10-fold cross validations to
optimzie the model's AUC (Area Under the Curve of the Receiver Operator
Characteristic). This resulting model was then tested on the 20\% of the
data that was originally held out from this overall process. Once the
ideal mtry was found the entire 490 sample set was used to create the
final Random Forest model on which testing on the 67-person cohort was
completed. The default cutoff of 0.5 was used as the threshold to
classify individuals as positive or negative for lesion.

\textbf{\emph{Selection of Important OTUs}} In order to assess which
variables were most central to all the models we counted the number of
times a variable was present in the top 10\% of mean decrease in
accuracy (MDA) for each different 80/20 split model and then filtered
this list to variables that were only present more than 50\% of the
time. This final collated list of variables was what was considered the
most important for the lesion prediction model.

A second model using the same types of conditions was used to identify
OTUs that could predict initial and follow up samples. These OTUs were
considered the most important for predicting initial and follow up
samples. This new model important OTUs were then compared to the overall
lesion model important OTUs and the resulting common OTUs to both models
were selected to test for changes between initial and follow up using a
paired wilcoxson rank sum test.

\textbf{\emph{Investigation of Important CRC Bacteria}} In order to
investigate how specific bacteria that have been previously associated
with CRC with selected OTUs that taxonomically classified to
Fusobacterium Nucleatum, Parvimonas Micra, Peptostreptococcus
Assacharolytica, and Porphyromonas Stomatis. Specifically we wanted to
observe if there were any differences based on whether the individual
had an adenoma or carcinoma that may have been missed with our lesion
versus normal model.

@@

\textbf{\emph{Reproducible methods.}} A detailed and reproducible
description of how the data were processed and analyzed can be found at
\url{https://github.com/SchlossLab/Baxter_followUps_2016}.

@@

>>>>

\textbf{\emph{Study Design and Patient Sampling}} The sampling and
design of the study was similar to that reported in Baxter, et al (13).
In brief, study exclusion involved those who had already undergone
surgery, radiation, or chemotherapy, had colorectal cancer before a
baseline stool sample could be obtained, had IBD, a known hereditary
non-polyposis colorectal cancer, or Familial adenomatous polyposis.
Samples used to build the model used for prediction were collected
either prior to a colonoscopy or between 1 - 2 weeks after. The
bacterial microbiome has been shown to nomralize within this time period
(\textbf{insert citation}). Kept apart from this training set were a
total of 67 individuals that not only had a sample as described
previoulsy but also a follow up sample between 188 - 546 days after
surgery and treatment had been completed. This study was approved by the
University of Michigan Institutional Review Board. All study
participants provided informed consent and the study itself conformed to
the guidelines set out by the Helsinki Declaration.

\textbf{\emph{Fecal Immunochemical Test and 16S rRNA Gene Sequencing}}
FIT was analyzed as previously published using both OC FIT-CHEK and
OC-Auto Micro 80 automated system (Polymedco Inc.) (\textbf{insert
citation}). 16S rRNA gene sequencing was completed as previously
described by Kozich, et al. (18). In brief, DNA extraction used the 96
well Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
automated pipetting system (Eppendorf). The V4 variable region was
amplified and the resulting product was split between three sequencing
runs with control, adenoma, and carcinoma evenly represented on each
run. Each group was randomly assigned to avoid biases based on sample
collection location.

@@

\textbf{\emph{Lesion Model Creation}} The Random Forest (19) algorithm
was used to create the model used for prediction of lesion (adenoma or
carcinoma) for the 67 individuals with follow up samples. The model
included data on FIT and the bacterial microbiome. Non-binary data was
checked for near zero variance and auto correlation. Data columns that
had near zero variance were removed. Columns that were correlated with
each other over a Spearman correlation coefficient of 0.75 had one of
the two columns removed. This pre-processing was performed with the R
package caret (v6.0.73). Optimization of the mtry hyperparameter
involved taking the samples and making 100 80/20 (train/test) splits in
the data where control and lesion were equally represented in the 80 and
20 split, respectively. This 80\% portion was then split again into an
80/20 split, and run through 20 repeated 10-fold cross validations to
optimzie the model's AUC (Area Under the Curve of the Receiver Operator
Characteristic). This resulting model was then tested on the 20\% of the
data that was originally held out from this overall process. Once the
ideal mtry was found the entire 490 sample set was used to create the
final Random Forest model on which testing on the 67-person cohort was
completed. The default cutoff of 0.5 was used as the threshold to
classify individuals as positive or negative for lesion. The
hyperparameter, mtry, defines the number of variables to investigate at
each split before a new division of the data is created.

\textbf{\emph{Initial Follow Up Model Creation}} We also investigated
whether a model could be created that could identify before and after
surgery samples. The training set utilized the 67-person cohort that was
previously used for testing of the lesion modeld. The creation of this
model and optimization of the mtry hyperparameter was completed using
the same procedure that was used to create the lesion model.

\textbf{\emph{Selection of Important OTUs}} In order to assess which
variables were most central to all the models we counted the number of
times a variable was present in the top 10\% of mean decrease in
accuracy (MDA) for each different 80/20 split model and then filtered
this list to variables that were only present more than 50\% of the
time. This final collated list of variables was what was considered the
most important for the lesion or initial follow up models.

@@

\textbf{\emph{Analysis Overview}} Differences in FIT between initial and
follow ups for either adenoma or carcinoma were investigate. Next,
initial and follow up samples were analyzed for differences in alpha and
beta diversity. All OTUs used in the lesion model were also analyzed
using a paired wilcoxson test. The lesion model was then tested for
accuracy in prediction and whether it reduced the positive probability
of lesion after surgery. The most important OTUs for this were used to
build an updated model and this reduced feature model was assessed for
it's similarity to the original model. We then used the initial follow
up model to assess whether this model could classify samples better then
the lesion model. The most important OTUs were then identified from this
model and used to create a reduced feature initial follow up model. This
reduced feature model, as was done with the lesion model, was compared
to the full model for loss of accuracy. Finally, in order to investigate
the relative abundance of specific bacteria, that have been previously
associated with CRC, we selected OTUs that taxonomically classified to
Fusobacterium Nucleatum, Parvimonas Micra, Peptostreptococcus
Assacharolytica, and Porphyromonas Stomatis. Specifically, we wanted to
test if there were any differences based on whether the individual had
an adenoma or carcinoma.

\textbf{\emph{Reproducible methods.}} A detailed and reproducible
description of how the data were processed and analyzed can be found at
\url{https://github.com/SchlossLab/Sze_followUps_2017}.

@@

\hypertarget{ref-kozich_development_2013}{}
18. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
dual-index sequencing strategy and curation pipeline for analyzing
amplicon sequence data on the MiSeq Illumina sequencing platform.
Applied and Environmental Microbiology \textbf{79}:5112--5120.
doi:\href{https://doi.org/10.1128/AEM.01043-13}{10.1128/AEM.01043-13}.

\hypertarget{ref-breiman_random_2001}{}
19. \textbf{Breiman L}. 2001. Random Forests. Machine Learning
\textbf{45}:5--32.
doi:\href{https://doi.org/10.1023/a:1010933404324}{10.1023/a:1010933404324}.
